---
title: "Multilevel_final"
output: pdf_document
---
## Multilevel Analysis

Since the pooled OLS and Fixed Effects Models do not account for the nested nature of the data and hence do not allow across-state analysis, Multilevel Coefficient Models (MCM) are applied in order to investigate the effects of individual and Bundesland-level factors on happiness. MCM examines individual variation within and across groups, as well as permits non-independence of individual factors from the group level variables. The first step of the MCM investigates whether statistically significant variation between Bundeslaender in terms of mean individual satisfaction exists. Otherwise, simpler OLS and panel-data models are more applicable for the given research. 

An unconditional model, *Null.Model*, serves the first purpose: the model only controls for the State that specifies that the variation intercept is a function of residence.

```{r, echo=FALSE}
Null.Model<-lme(satis~1,random=~1|Stateid,data=data,
                  control=list(opt="optim"))
VarCorr(Null.Model)
``` 

The model examines how much of the average individual life satisfaction is explained by the residence of a respondent through R's general purpose optimization routine (opt="optim"). According to the *Null.Model*, the Bundesland variation (intercept variance) is 0.19, while the within-State residual is 3.17. 

```{r, echo=FALSE}
GREL.DAT<-GmeanRel(Null.Model)
names(GREL.DAT)
mean(GREL.DAT$MeanRel)
Null.Model.2<-gls(satis~1,data=data,
                    control=list(opt="optim"))
logLik(Null.Model.2)*-2
logLik(Null.Model)*-2
sigtest <- round((logLik(Null.Model.2)*-2)-(logLik(Null.Model)*-2), digits=3)
```

Furthermore, the *GmeanRel* function (*multilevel* package) yields the mean reliability of 16 Bundeslaender, which, in this case, is substantially high (`r round(mean(GREL.DAT$MeanRel), digits=3)`). As the cut-off point for acceptable reliability is 0.7, Bindeslaeander group reliability meets the threshold. The difference between the -2 log likelihood values of *Null.Model* and *Null.Model.2* tests whether the between-State effect is present compared to the random variation in life satisfaction without any control variables. According to the results, the difference is significantly large based on the Chi-Squared distribution with 2 degrees of freedom (`r sigtest`). These results suggest that there is significant State variation in happiness level, which justifies the usage of the MCM. 

The second step of the MCM looks into how both Bundesland emissions and subjective concerns about the environment influence reported individual life satisfaction. *Model.1* (*lme* function) for the time being does not control for individual characteristics, such as gender, age, employment and family status. Moreover, percentage of the renewables (which is anticipated to directly relate to the life satisfaction) will be also examined in the later phase of the research. *Model.1* serves as the basic tool to understand relationship between the group- and individual-level factors representing the environment. 

```{r, echo=FALSE}
Model.1<-lme(satis~environ+Emissions,random=~1|Stateid,data=finaldata,
             control=list(opt="optim"))

summary(Model.1)
```

The output demonstrates that the State emissions are indeed negatively related to the individual happiness. However, after controlling for the individual environmental concerns, the coefficient (-0.005) is not statistically significant. Consequently, the group-level effect does not significantly differ from the individual-level. On the other hand, the coefficient of environmental concerns statistically differs from 0, but the direction is oddly positive counter the expectations. These results point at the omitted variables (absence of the demographic characteristics) that should be included in the further model. Similarly, the surprising direction of the environmental coefficient demands investigation of endogeneity between reported happiness and environmental concerns. 

A brief overview of *Model.2*, which covers demographic characteristics, reassures the anticipation about the State emissions: the coefficient became significant after controlling for other individual factors. As a result, the difference between State- and individual-level slopes exists and permit multilevel analysis. However, the coefficient of environmental concerns is still positive, which is to be discussed together with other factors in the final paper. 

```{r, echo=FALSE}
Model.2<-lme(satis~environ+Emissions+age+fam+gender+emp,random=~1|Stateid,data=finaldata,
             control=list(opt="optim"))

summary(Model.2)
```