## Emissions Levels and Life Satisfaction
### A Study on the German States

```{r echo=FALSE, results='hide', message=FALSE}
library(DataCombine)
library(shiny)

suppressPackageStartupMessages(library(googleVis))

#load dataset
data <- read.csv("All_Merged_Data.csv")
attach(data)

## Round Emissions
transform(data, Emissions = is.numeric(Emissions))
data <- transform(data,  Emissions=round (Emissions,1))

#aggregate yearly emissions data by state
agg <- aggregate(cbind(CO2perSqKm) ~ State + Year + Stateid, FUN=mean, data=data)

allvar <- aggregate(cbind(CO2perSqKm, satis, environ, GrossIncome) ~ State + Year + Stateid, FUN=mean, data=data)

#Transform yearly data to numeric
transform(agg, Year = as.numeric(Year))

Replace1 <- data.frame(from = c("Baden-W\xcc_rttemberg", "Th\xcc_ringen"), to = c("Baden-Wuerttemberg", "Thueringen"))
                       
allvar <- FindReplace(data=allvar, Var = "Stateid", replaceData = Replace1, from = "from", to = "to",
  exact = TRUE, vector = FALSE)

allvar <- FindReplace(data=allvar, Var = "State", replaceData = Replace1, from = "from", to = "to",
  exact = TRUE, vector = FALSE)

#create new vector with state codes

Replaces <- data.frame(from = c("Baden-Wuerttemberg", "Bayern", "Bremen", "Berlin", "Brandenburg", "Hamburg" , "Hessen", "Mecklenburg-Vorpommern", "Niedersachsen", "Nordrhein-Westfalen", "Rheinland-Pfalz", "Sachsen", "Sachsen-Anhalt", "Schleswig-Holstein", "Saarland", "Thueringen"), to = c("DE-BW", "DE-BY", "DE-HB", "DE-BE", "DE-BB", "DE-HH", "DE-HE", "DE-MV", "DE-NI", "DE-NW", "DE-RP", "DE-SN", "DE-ST", "DE-SH", "DE-SL", "DE-TH"))

New <- FindReplace(data=allvar, Var = "Stateid", replaceData = Replaces, from = "from", to = "to",
  exact = TRUE, vector = FALSE)

New[, c(2,4)] <- sapply(New[, c(2,4)], as.numeric)


## Create Geo Charts for each year

geoChartDE <- list(region="DE", 
                   resolution="provinces",
                   legend="{numberFormat:'#,###.00'}") 
```

## Introduction
Climate change is undoubtedly a global problem, but this fact means it is also, in a way, a classic tragedy of the commons. No country wants to put itself at an economic disadvantage by restricting the use of cheap fossil fuels so all continue to emit, deteriorating the "commons" of the Earth's protective atmosphere. This can make it difficult to get citizens to identify with the problem and take responsibility; people will more likely act when something affects them individually. In this paper we want to explore this aspect: how much do people feel, consciously or unconsciously, the effects of green house gas emissions? More specifically, do emissions affect their reported health, well-being, or life satisfaction?  

To examine this question, we will look at data from Germany. Germany is a leader in protecting the environment while also having a long history as an industrial power and coal producer. One one hand, its energy transition (*Energiewende*) is considered one of the most ambitious climate policy projects in the world. On the other hand, it has struggled with appropriate incentives, a drop in oil prices---not to mention coal's continued role as a cheap and reliable fuel---and how to transform the transportation sector. Germany therefore still does emit large amounts of green house gases. We will look at green house gas emissions data by federal state (*Bundesland*) and compare that with life satisfaction data to examine our first hypothesis:

> *H1: Bundeslaender with higher emissions will have lower reported levels of health, well-being, or life satisfaction.* 

We will also investigate whether there is a time component to perceptions of life satisfaction and well-being. This is important because, as explained in more depth in the literature review, there is a possibe endogeneity problem. Individuals may be unhappy with more pollution, but unhappy individuals may be bothered more by pollution. Germany's emissions have been reduced since 1990, though reductions have stagnated recently. By tracking responses over time, we hope to tell if there is a time component: do changes in emissions precede changes in life satisfaction? There are some studies that have used a time series approach, but not many have done so a a country level. Using German emissions and happiness data going back to 1990, our second hypothesis is:

> *H2: Reported levels of health, well-being or life satisfaction will reflect changes in emissions in line with hypothesis above, i.e. as emissions decline, reported levels of health, well-being, and life satisfaction will rise.* 

## Average Emissions per Square KM by State
```{r, echo=FALSE, results='asis'}
# Plot
Map <- gvisGeoChart(New, locationvar = 'State',
                      colorvar = 'CO2perSqKm',
                      options = geoChartDE)

plot(Map, tag = 'chart')



```


## Comparing Effects of Variables on Life Satisfaction by State
The dynamic chart below allows an examination of the correlation between the variables of interest and overall life satisfaction. The highlights the changes for each of the German states between the years 1990-2012. Scroll through the years at the bottom to see the changes in values over time for each state.

```{r, echo=FALSE, results='asis'}
M1 <- gvisMotionChart(allvar, idvar="State", timevar="Year")
plot(M1, tag = 'chart')

```


## Testing running code

```{r, echo=FALSE}

#Boxplot of emissions levels by state

boxplot(Emissions~State,
  data=data,
  main="Emissions by State",
  xlab="State",
  ylab="Emissions (annual tons of CO2 per capita)",
  col="royalblue1",
  border="royalblue4"
)
```